---
title: "About the Data"
---

## Data Exploration:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(dplyr)
library(plotly)
suppressMessages(source("raw_data.R"))

```

**What are the data dimensions?**

```{r, echo=FALSE, warning=FALSE}
dum<-dim(merged_data)
cat("There are 19 columns and 15674 observations.")
```

**How many families reported their data?**

```{r, echo=FALSE, warning=FALSE}
#Filtered HHMEM==1 
one_person_households <- merged_data %>%
  filter(n_hhmem == 1)
count_one_person_households <- nrow(one_person_households)

cat("There are", count_one_person_households, "reported households that have 1 person.")
```

**What is the average household size?**

```{r, echo=FALSE, warning=FALSE}

household <- merged_data %>%
  group_by(hhid_a) %>%
  summarize(max_n_hnmem = max(n_hhmem, na.rm = TRUE))
average_max_n_hnmem <- mean(household$max_n_hnmem, na.rm = TRUE)
cat("There average house hold size is", average_max_n_hnmem,"people.")

```

**What is the youngest age?**

```{r, echo=FALSE, warning=FALSE}
min_age<-min(merged_data$age_yrs, na.rm = TRUE)
cat("The youngest person in the dataset is", min_age, "years old.")
```

**What is the oldest age?**

```{r, echo=FALSE, warning=FALSE}
max_age<-max(merged_data$age_yrs, na.rm = TRUE)
cat("The oldest person in the dataset is", max_age, "years old.")
```

The gender distribution indicates that 43.3% of respondents identify as female, 52.4% as male, and 4.3% did not provide a response.

**What is the race distribution?**

```{r, echo=FALSE, warning=FALSE}



category_counts <- merged_data %>%
  filter(!is.na(race_nofa)) %>%
  group_by(race_nofa) %>%
  summarise(count = n())
total <- sum(category_counts$count)

category_counts <- category_counts %>%
  mutate(proportion = (count / total) * 100)

pie_chart <- plot_ly(category_counts, labels = ~factor(race_nofa), values = ~proportion, type = 'pie', hole = 0.6) %>%
  layout(title = "Proportion of Races", showlegend = TRUE)
pie_chart

```

For those who reported their race, 48% of the surveyed individuals are White. Similarly, 44.8% of the survey individuals are Black. About 3.7% of the data are represented by other racial groups.

**What is the gender distribution?**

```{r, echo=FALSE, warning=FALSE}

# Assuming 'merged_data' is your data frame
merged_data$male <- as.factor(merged_data$male)
gender_counts <- table(merged_data$male)
gender_counts_df <- as.data.frame(gender_counts)
names(gender_counts_df) <- c("Gender", "Count")

library(ggplot2)

ggplot(gender_counts_df, aes(x = Gender, y = Count / sum(Count), fill = Gender)) +
  geom_col(color = "black") +
  geom_text(aes(label = scales::percent(Count / sum(Count), accuracy = 0.1)),
            position = position_stack(vjust = 0.5), size = 4, color = "black") +
  labs(title = "Distribution of Gender", x = "Gender", y = "Proportion") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  ) +
  scale_fill_manual(values = c("lightgreen", "#fc8d62"), labels = c("Female", "Male")) +
  scale_x_discrete(labels = c("Female", "Male")) +
  labs(fill = "Legend")

```

Among individuals who provided information about their gender, approximately 54% identified as male, while 45.3% identified as female.

**What is the distribution of sexual orientation?**

```{r, echo=FALSE, warning=FALSE}

# Assuming 'merged_data' is your data frame
merged_data$straight <- as.factor(merged_data$straight)
merged_data$gay <- as.factor(merged_data$gay)
merged_data$bisexual <- as.factor(merged_data$bisexual)

sexual_orientation_percentages <- data.frame(
  SexualOrientation = c("Straight", "Gay", "Bisexual"),
  Percentage = c(
    sum(merged_data$straight == "Yes", na.rm = TRUE) / nrow(merged_data) * 100,
    sum(merged_data$gay == "Yes", na.rm = TRUE) / nrow(merged_data) * 100,
    sum(merged_data$bisexual == "Yes", na.rm = TRUE) / nrow(merged_data) * 100
  )
)

library(ggplot2)

# Reorder the bars from highest to lowest percentage
sexual_orientation_percentages$SexualOrientation <- factor(sexual_orientation_percentages$SexualOrientation,
                                                            levels = sexual_orientation_percentages$SexualOrientation[order(-sexual_orientation_percentages$Percentage)])

ggplot(sexual_orientation_percentages, aes(x = SexualOrientation, y = Percentage, fill = SexualOrientation)) +
  geom_col(color = "black") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")),
            position = position_stack(vjust = 0.5), size = 4, color = "black") +
  labs(title = "Percentage Distribution of Sexual Orientation", x = "Sexual Orientation", y = "Percentage") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold")
  ) +
  scale_fill_manual(values = c("#66c2a5", "#fc8d62", "#8da0cb"),
                    labels = c("Straight", "Gay", "Bisexual")) +
  labs(fill = "Legend")

```

Among those who disclosed their sexual orientation, 61% identified as straight, 2.1% as gay, and 1.9% as bisexual.

**What is the proportion of substance type?**

```{r, echo=FALSE, warning=FALSE}
merged_data$subsAbuse_type <- as.factor(merged_data$subsAbuse_type)

merged_data_filtered <- merged_data[!is.na(merged_data$subsAbuse_type), ]

subsAbuse_percentages <- prop.table(table(merged_data_filtered$subsAbuse_type)) * 100

subsAbuse_df <- as.data.frame(subsAbuse_percentages)
names(subsAbuse_df) <- c("SubstanceAbuse", "Percentage")

subsAbuse_df <- subsAbuse_df[order(-subsAbuse_df$Percentage), ]

ggplot(subsAbuse_df, aes(x = reorder(SubstanceAbuse, -Percentage), y = Percentage, fill = SubstanceAbuse)) +
  geom_col(color = "black") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")),
            position = position_stack(vjust = 0.5), size = 4, color = "black") +
  labs(title = "Percentage Distribution of Substance Abuse Types", x = "Substance Abuse Type", y = "Percentage") +
  theme_minimal() +
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 14, face = "bold"), plot.title = element_text(size = 16, face = "bold")) +
  scale_fill_manual(values = c("lightgreen", "#8da0cb", "darkgrey", "#fc8d62", "lightyellow"),
                    labels = levels(merged_data_filtered$subsAbuse_type)) +
  labs(fill = "Legend")


```

Approximately 79.8% of the individuals surveyed indicated that they refrain from substance abuse. About 8.4% reported engaging in the abuse of both drugs and alcohol, while 4.6% experience alcohol abuse and 7.2% experience drug abuse.
